\Large\textbf{Advection versus diffusion in brain ventricular transport}

### 2025-04-13 paper/main.tex
add zenodo repo link, align figures

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b, c.} Particle tracking data visualizing CSF flow directionality (\textbf{b}) and
    velocity magnitudes (\textbf{c}) in the middle ventricle.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue). Dorsal (D), ventral (V), anterior (A), posterior (P),
    left (L) and right (R) coordinates indicated.
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-driven/no-cardiac flow model.
    \textbf{f.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries. Time instants of streamline plots indicated.
    \textbf{g.} Streamlines of the CSF flow field simulated with the cardiac-induced/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{h--j.} Streamlines simulated with the baseline flow model (cilia+cardiac)
    at a time 25\% (\textbf{h}), 50\% (\textbf{i}) and 75\% (\textbf{j})
    into the cardiac cycle. %t=0.112613 s %t=0.225225 s %t=0.315315 s
    Scale bars 50 \textmu m.
    }
    \label{fig:fig1}
\end{figure}

@@

\begin{figure}%[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure2_simulation_results_different_coeffs.png}
    \caption{Simulated distribution and evolution after photoconversion of
    extracellular vesicles (EVs, $D_1$), STM-GFP ($D_2$) and Dendra2 ($D_3$). 
    \textbf{a.} Geometry-centered clip of the ventricles mesh showing the locations of the 
    regions of interest (ROIs).
    \textbf{b--d.} Slices of the concentration field simulated
    with diffusion coefficients $D_1$, $D_2$ and $D_3$ for nine time instants covering the
    time range from 90 seconds to 811 seconds, corresponding to 200 to 1800 cardiac cycles,
    for even intervals of 200 cycles.
    Slices are taken in the $xz$-plane ($y=0.140$ mm, the center of the geometry).
    \textbf{e--g.} Mean concentration $\cbar$ in each ROI
    as function of time for diffusion coefficients $D_1$, $D_2$ and $D_3$. 
    \textbf{h.} The time $\hat{t}$ before the mean concentration $\cbar$ exceeds a
    threshold value of $0.25$ (ROIs 2-4) or $0.10$ (ROIs 5, 6) in each simulation setup.
    Note that there is no bar for $D_1$ in ROI 6, because the threshold value was never exceeded.}
    \label{fig:fig2}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.975\textwidth]{graphics/figure3_compare_experiments_and_simulations.png}
    \caption{
    \textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$ 
    in each region of interest (ROI)
    observed during experiments monitoring
    the movement of Dendra2 proteins photoconverted by laser targeting ROI 1. 
    Bold lines are mean values of the cohort ($n=8$) and shaded regions cover
    one standard deviation errors in the mean.
    \textbf{b.} Mean concentration $\cbar(t)$ in each ROI in a
    simulation of Dendra2 protein transport with
    the diffusion coefficient $D_3$.
    \textbf{c.} Time to $\Delta F(t)$ reaches a threshold value of 0.25 in
    one representative zebrafish larva.
    \textbf{d.} Time to $\cbar(t)$ reaches a threshold value of 0.25 for
    a transport simulation with diffusion coefficient $D_3$.
    We have excluded the posterior-most regions of the geometries,
    because the values of $\Delta F$ and $\cbar$ never exceeded the threshold
    value of 0.25 in these regions. Scale bars 50 \textmu m.}
    \label{fig:fig3}
\end{figure}
% Final Delta F values
%0.6322650156762044 ROI 2
%0.7059449124175479 ROI 3
%0.6694500367160006 ROI 4
%0.28986700094045437 ROI 5
%0.18155055624229066 ROI 6
% Final mean c values ROIs 2-6: 0.80, 0.76, 0.86, 0.60 and 0.38

@@

In the computational model, we represent the \emph{smh} mutants by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}). Compared to experimental
data, the control and mutant scenarios differed more for the
simulations, with the times-to-threshold in ROI 2 and 4 being 81\% and
28\% higher in the absence of cilia-mediated flow. Moreover, the
predicted mean concentrations in ROI 5 deviate between the two
scenarios, while in the experiments the difference in
fluorescence intensities is not significant.
Altogether, our findings identify that,
for smaller proteins such as Dendra2,
diffusion plays an important role for solute movement to regions
upstream in the cilia-mediated flow fields or to distant ventricles,
while the cilia affect the downstream mean concentrations.
\begin{figure}[H]
    \centering
    \includegraphics[width=0.96\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$
    in each region of interest (ROI) for control (black lines)
    and \emph{smh} mutant (red lines) zebrafish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard deviation.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for controls (gray dots) and
    \emph{smh} mutants (red dots).
    The thick black lines denote mean values. The $p$-values
    were calculated using a Mann-Whitney U test with 95\% confidence level,
    under the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-induced/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in panels \textbf{a} and \textbf{c} the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

Turning to transport of Dendra2, removal of the
dorsal cilia again has the most pronounced effect. All mean concentration
profiles (except ROI 1, where we impose the photoconversion curve)
change noticeably (\Crefpanels{fig:fig5}{c}). Compared to the baseline
model results, we observe less solute distribution in the middle and
anterior ventricles (ROIs 2--5), and more rapid transport to the
posterior ventricle (ROI 6). On the other hand, removal of anterior
cilia does not alter the mean concentrations
(\Crefpanels{fig:fig5}{c}). Finally, when removing ventral cilia in the
middle ventricle more solute spreads towards the anterior ventricle
(ROI 5), and less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate cilia regions for which cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} The total tangential force applied to the boundary with the tangential traction
    boundary condition in each of the three separate ciliated regions, as well as the total force
    (the sum of the three regions), calculated with \Cref{eq:tangential_force_calculation}. 
    \textbf{c.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{\textbf{a.} The original computational mesh of the ventricular geometry.
    \textbf{b.} The four modified versions of the computational geometry. Compared
    to the original mesh, the volumes of the deformed meshes are reduced by 3.9\%, 4.1\%, 1.4\% 
    and 9.4\%.
    \textbf{c.} Mean concentration $\cbar(t)$ simulated with the original and modified geometries
    in the six regions of interest (ROIs).
    \textbf{d.} A comparison of the tangential cilia force in the three cilia regions
    and the total force applied, calculated with \Cref{eq:tangential_force_calculation}.
    In each region, the force $F_{\mathrm{original}}$ 
    applied on the original mesh is the baseline of 100\%,
    to which the forces $F_i$ of the other mesh versions are compared, where $i$ denotes
    the modification scenario.
    The bars for the four alternative scenarios display $F_i/F_{\mathrm{original}}\times 100\%$.}
    \label{fig:fig6}
\end{figure}

@@

@@

\subsection{Photoconversion data analysis}
The image data analysis was performed in MATLAB and the codes are
openly available~\fixme{\cite{zenodo-link-to-all-code}}.  The
photoconverted channel was aligned to correct for drift in $x,y$
directions using previously developed
codes~\cite{Reiten2017Motile-Cilia-MediatedComputations,
  Ringers2023NovelEpithelia}.  Only stable recordings were further
analyzed.  To increase the signal-to-noise ratio, the images were
downsampled by a block average using a resampling factor of 6. 
The change in fluorescence intensity
$\Delta F = (F-F_0)/F_0$ was
calculated for each pixel and each time point.  Here, the
value $F_0$ is the average fluorescence intensity for the baseline
(the first 10 time-lapse images) acquired before photoconversion.

@@

The numerical solution schemes for the Stokes equations and the
advection-diffusion equation were implemented and solved numerically with
DOLFINx with petsc4py as linear algebra backend
\cite{dolfinx2023preprint, Dalcin2011ParallelPython}.
The simulation code is openly
available\fixme{~\cite{zenodo-link-to-all-code}}.
A direct solver using MUMPS~\cite{Amestoy2011Mumps} is used to solve the
linear systems resulting from the discretized Stokes equations.
For the discretized advection-diffusion equation, we use a flexible GMRES solver
with block Jacobi preconditioning \cite{fgmres, Jacobi1845UeberGleichungen}. 
We discretize the time-dependent diffusion equation (without advection)
and use the assembled linear system matrix
as the preconditioner matrix for the advection-diffusion problem.
Numerical experiments verified the accuracy of the model implementation
(\Cref{sec:appB_verification}).

@@

% Data Availability Statement
\medskip
\noindent \textbf{Data Availability Statement} \par
\noindent The data and model implementations presented in this paper are
openly available at \fixme{REPO-LINK}.

@@

Particle movement varied over the different measurements. The mean of $\overline{u}_x$ was
0.26 \textmu m/s and the mean of $\overline{u}_z$ was 0.08 \textmu m/s.
The maximum mean velocities
in absolute values were 10.2 \textmu m/s and 2.4 \textmu m/s for
$\overline{u}_x$ and $\overline{u}_z$,
respectively. We used the cardiac-induced/no-cilia flow model to tune the
pressure boundary condition
amplitude $A$. The choice of $A=1.5$ mPa was used since: (i) it yields a
maximum velocity magnitude
of 4.8 \textmu m/s, which is close to the maximum observed in the
particle tracking data when
rejecting the outlier at around 10 \textmu m/s; (ii) it yields a time-averaged mean velocity
of 0.89 \textmu m/s, which is close to the mean observed in the particle tracking data.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure7.png}
    \caption{Particle tracking data. Mean velocities in
    the $x$ (\textbf{a}) and $z$ (\textbf{b})
    directions of the particles tracked.
    The dashed horizontal lines designate zero.}\label{fig:particle_tracking_data}
\end{figure}

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b, c.} Particle tracking data visualizing CSF flow directionality (\textbf{b}) and
    velocity magnitudes (\textbf{c}) in the middle ventricle.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue). Dorsal (D), ventral (V), anterior (A), posterior (P),
    left (L) and right (R) coordinates indicated.
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-driven/no-cardiac flow model.
    \textbf{f.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries. Time instants of streamline plots indicated.
    \textbf{g.} Streamlines of the CSF flow field simulated with the cardiac-induced/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{h--j.} Streamlines simulated with the baseline flow model (cilia+cardiac)
    at a time 25\% (\textbf{h}), 50\% (\textbf{i}) and 75\% (\textbf{j})
    into the cardiac cycle. %t=0.112613 s %t=0.225225 s %t=0.315315 s
    Scale bars 50 \textmu m.
    }
    \label{fig:fig1}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{graphics/figure2_simulation_results_different_coeffs.png}
    \caption{Simulated distribution and evolution after photoconversion of
    extracellular vesicles (EVs, $D_1$), STM-GFP ($D_2$) and Dendra2 ($D_3$). 
    \textbf{a.} Geometry-centered clip of the ventricles mesh showing the locations of the 
    regions of interest (ROIs).
    \textbf{b--d.} Slices of the concentration field simulated
    with diffusion coefficients $D_1$, $D_2$ and $D_3$ for nine time instants covering the
    time range from 90 seconds to 811 seconds, corresponding to 200 to 1800 cardiac cycles,
    for even intervals of 200 cycles.
    Slices are taken in the $xz$-plane ($y=0.140$ mm, the center of the geometry).
    \textbf{e--g.} Mean concentration $\cbar$ in each ROI
    as function of time for diffusion coefficients $D_1$, $D_2$ and $D_3$. 
    \textbf{h.} The time $\hat{t}$ before the mean concentration $\cbar$ exceeds a
    threshold value of $0.25$ (ROIs 2-4) or $0.10$ (ROIs 5, 6) in each simulation setup.
    Note that there is no bar for $D_1$ in ROI 6, because the threshold value was never exceeded.}
    \label{fig:fig2}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure3_compare_experiments_and_simulations.png}
    \caption{
    \textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$ 
    in each region of interest (ROI)
    observed during experiments monitoring
    the movement of Dendra2 proteins photoconverted by laser targeting ROI 1. 
    Bold lines are mean values of the cohort ($n=8$) and shaded regions cover
    one standard deviation.
    \textbf{b.} Mean concentration $\cbar(t)$ in each ROI in a
    simulation of Dendra2 transport with
    the diffusion coefficient $D_3$.
    \textbf{c.} Time to $\Delta F(t)$ reaches a threshold value of 0.25 in
    one representative zebrafish larva.
    \textbf{d.} Time to $\cbar(t)$ reaches a threshold value of 0.25 for
    a transport simulation with diffusion coefficient $D_3$.
    The posterior-most regions of the geometries are excluded,
    because $\Delta F$ and $\cbar$ never exceeded the threshold
    value of 0.25 in these regions. Scale bars 50 \textmu m.}
    \label{fig:fig3}
\end{figure}
% Final Delta F values
%0.6322650156762044 ROI 2
%0.7059449124175479 ROI 3
%0.6694500367160006 ROI 4
%0.28986700094045437 ROI 5
%0.18155055624229066 ROI 6
% Final mean c values ROIs 2-6: 0.80, 0.76, 0.86, 0.60 and 0.38

@@

In the computational model, we represent the \emph{smh} mutants by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}). Compared to experimental
data, the control and mutant scenarios differed more for the
simulations, with the times-to-threshold in ROI 2 and 4 being 81\% and
28\% higher in the absence of cilia-mediated flow. Moreover, the
predicted mean concentrations in ROI 5 deviate between the two
scenarios, while in the experiments the difference in
fluorescence intensities is not significant.
Altogether, our findings identify that,
for smaller proteins such as Dendra2,
diffusion plays an important role for solute movement to regions
upstream in the cilia-mediated flow fields or to distant ventricles,
while the cilia affect the downstream mean concentrations.
\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$
    in each region of interest (ROI) for control (black lines)
    and \emph{smh} mutant (red lines) zebrafish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard deviation.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for controls (gray dots) and
    \emph{smh} mutants (red dots).
    The thick black lines denote mean values. The $p$-values
    were calculated using a Mann-Whitney U test with 95\% confidence level,
    under the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-induced/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in panels \textbf{a} and \textbf{c} the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

Turning to transport of Dendra2, removal of the
dorsal cilia again has the most pronounced effect. All mean concentration
profiles (except ROI 1, where we impose the photoconversion curve)
change noticeably (\Crefpanels{fig:fig5}{c}). Compared to the baseline
model results, we observe less solute distribution in the middle and
anterior ventricles (ROIs 2--5), and more rapid transport to the
posterior ventricle (ROI 6). On the other hand, removal of anterior
cilia does not alter the mean concentrations
(\Crefpanels{fig:fig5}{c}). Finally, when removing ventral cilia in the
middle ventricle more solute spreads towards the anterior ventricle
(ROI 5), and less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate cilia regions for which cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} The total tangential force applied to the boundary with the tangential traction
    boundary condition in each of the three separate ciliated regions, as well as the total force
    (the sum of the three regions), calculated with \Cref{eq:tangential_force_calculation}. 
    \textbf{c.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

@@

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{\textbf{a.} The original computational mesh of the ventricular geometry.
    \textbf{b.} The four modified versions of the computational geometry. Compared
    to the original mesh, the volumes of the deformed meshes are reduced by 3.9\%, 4.1\%, 1.4\% 
    and 9.4\%.
    \textbf{c.} Mean concentration $\cbar(t)$ simulated with the original and modified geometries
    in the six regions of interest (ROIs).
    \textbf{d.} A comparison of the tangential cilia force in the three cilia regions
    and the total force applied, calculated with \Cref{eq:tangential_force_calculation}.
    In each region, the force $F_{\mathrm{original}}$ 
    applied on the original mesh is the baseline of 100\%,
    to which the forces $F_i$ of the other mesh versions are compared, where $i$ denotes
    the modification scenario.
    The bars for the four alternative scenarios display $F_i/F_{\mathrm{original}}\times 100\%$.}
    \label{fig:fig6}
\end{figure}

@@

\subsection{Photoconversion data analysis}
The image data analysis was performed in MATLAB and the codes are
openly available~\cite{Herlyng2025Zenodo}.  The
photoconverted channel was aligned to correct for drift in $x,y$
directions using previously developed
codes~\cite{Reiten2017Motile-Cilia-MediatedComputations,
  Ringers2023NovelEpithelia}.  Only stable recordings were further
analyzed.  To increase the signal-to-noise ratio, the images were
downsampled by a block average using a resampling factor of 6. 
The change in fluorescence intensity
$\Delta F = (F-F_0)/F_0$ was
calculated for each pixel and each time point.  Here, the
value $F_0$ is the average fluorescence intensity for the baseline
(the first 10 time-lapse images) acquired before photoconversion.

@@

The numerical solution schemes for the Stokes equations and the
advection-diffusion equation were implemented and solved numerically with
DOLFINx with petsc4py as linear algebra backend
\cite{dolfinx2023preprint, Dalcin2011ParallelPython}.
The simulation code is openly
available~\cite{Herlyng2025Zenodo}.
A direct solver using MUMPS~\cite{Amestoy2011Mumps} is used to solve the
linear systems resulting from the discretized Stokes equations.
For the discretized advection-diffusion equation, we use a flexible GMRES solver
with block Jacobi preconditioning \cite{fgmres, Jacobi1845UeberGleichungen}. 
We discretize the time-dependent diffusion equation (without advection)
and use the assembled linear system matrix
as the preconditioner matrix for the advection-diffusion problem.
Numerical experiments verified the accuracy of the model implementation
(\Cref{sec:appB_verification}).

@@

% Data Availability Statement
\medskip
\noindent \textbf{Data Availability Statement} \par
\noindent The software and data used and presented in this paper are
openly available\cite{Herlyng2025Zenodo}.

@@

Particle movement varied over the different measurements. The mean of $\overline{u}_x$ was
0.26 \textmu m/s and the mean of $\overline{u}_z$ was 0.08 \textmu m/s.
The maximum mean velocities
in absolute values were 10.2 \textmu m/s and 2.4 \textmu m/s for
$\overline{u}_x$ and $\overline{u}_z$,
respectively. We used the cardiac-induced/no-cilia flow model to tune the
pressure boundary condition
amplitude $A$. The choice of $A=1.5$ mPa was used since: (i) it yields a
maximum velocity magnitude
of 4.8 \textmu m/s, which is close to the maximum observed in the
particle tracking data when
rejecting the outlier at around 10 \textmu m/s; (ii) it yields a time-averaged mean velocity
of 0.89 \textmu m/s, which is close to the mean observed in the particle tracking data.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure7.png}
    \caption{Particle tracking data. Mean velocities in
    along the rostrocaudal axis ($x$ axis, \textbf{a})
    and along the dorsoventral axis ($z$ axis, \textbf{b})
    directions of the particles tracked.
    The dashed horizontal lines designate zero.}\label{fig:particle_tracking_data}
\end{figure}
